---
title: "Data Visualisation with ggplot2 - Project Assignment"
output:
  html_document:
    df_print: paged
---

# AfterWork Data Science: Data Visualisation with Python Project

## 1. Defining the Question

### a) Specifying the Data Analysis Question
Specify the reasearch question that you'll be answering. i.e. Provide strategy recommendations that will lead to revenue growth.

### b) Defining the Metric for Success
The solutions to the following questions will help us answer our research question:

* When is the best time of year to book a hotel room?
* When is the optimal length of stay in order to get the best daily rate?
* How will you know if a hotel was likely to receive a disproportionately high number of special requests?

### c) Understanding the context
Provide some background information....

### d) Recording the Experimental Design
Describe the steps/approach that you will use to answer the given question.

### e) Data Relevance
How relevant was the provided data?


## 2. Reading the Data
```{r}
library("tidyverse")
```

```{r}
# Load the data below
# --- 
# Dataset url = 
# --- 
# YOUR CODE GOES BELOW
# 

hotel_df <- read_csv("hotel_bookings.csv")
head(hotel_df)
```

```{r}
# Checking the first 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 

head(hotel_df,5)
```

```{r}
# Checking the last 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
tail(hotel_df,5)
```

```{r}
# Sample 10 rows of data
# ---
# YOUR CODE GOES BELOW
# 
sample_n(hotel_df,10)
```

```{r}
# Checking number of rows and columns
# ---
# YOUR CODE GOES BELOW
#  

dim(hotel_df)
```

```{r}
# Checking datatypes
# ---
# YOUR CODE GOES BELOW
# 
str(hotel_df)
```

```{r}
glimpse(hotel_df)
```

Record your general observations below:

Observation 1
  We have 119390 rows and 32 columns of data in our dataset
Observation 2
  

## 3. External Data Source Validation
The data is originally from the article Hotel Booking Demand Datasets, by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.


## 4. Data Preparation
### Performing Data Cleaning
```{r}
# Checking datatypes and missing entries of all the variables
# ---
# YOUR CODE GOES BELOW
# 
colSums(is.na(hotel_df))
```

We observe the following from our dataset:

Observation 1
  We only have column children missing only 4 records
Observation 2

```{r}
# Checking how many duplicate rows are there in the data
# ---
# YOUR CODE GOES BELOW
# 
hotel_df[duplicated(hotel_df),]
```
```{r}
unique(hotel_df)
```

We choose to keep the duplicates because we don't have a unique identifier to actually proof that we have duplicates.

```{r}
# Checking if any of the columns are all null
# ---
# YOUR CODE GOES BELOW
# 

colSums(is.na(hotel_df))
```

We observe the following from our dataset:

Observation 1
  We do not have any column with all fields missing

```{r}
# Checking if any of the rows are all null
# ---
# YOUR CODE GOES BELOW
# 
rowSums(is.na(hotel_df))
```

We observe the following from our dataset:

Observation 1
  We have alot of data missing 118390
Observation 2

```{r}
# Checking the correlation of the features through the use of 
# visualizations the correlation using heatmap
# ---
# YOUR CODE GOES BELOW
# 

install.packages("reshape2")

# loading the package
library(reshape2)
```

We observe the following from our dataset:

Observation 1
Observation 2

```{r}
# Dropping company column because it has alot of missing values 
# and we won't need to answer any of our questions
# ---
# YOUR CODE GOES BELOW
# 

hotel1_df = select(hotel_df, -c("company"))
head(hotel1_df)
```

From the data variable description we see that the Distribution Channel categoricy that tells us about Booking distribution.

The term “TA” means “Travel Agents”
The term “TO” means “Tour Operators”
This allows us to fill the missing values in the agents column with TO

```{r}
# We replace the mising values i.e. for TO
# ---
# YOUR GOES BELOW
#

hotel1_df$agent[is.na(hotel1_df$agent)] <- "TO"
hotel1_df
```

```{r}
# We drop rows where there is no adult, baby and child as 
# these records won't help us.
# ---
# YOUR GOES BELOW
#

hotel1_df %>%
  filter(adults>0,children>0,babies>0)
```

```{r}
# We replace missing children values with rounded mean value
# ---
# Hint i.e. use round()
# ---
# YOUR GOES BELOW
#

hotel1_df$children[is.na(hotel1_df$children)] <- round(mean(hotel1_df$children, na.rm = TRUE))
hotel1_df
```

```{r}
# Checking for missing values in the dataframe
# ---
# YOUR GOES BELOW
#
colSums(is.na(hotel1_df))
```

```{r}
# Converting the datatypes of the following columns from float to integer
# i.e. children, company, agent
# ---
# YOUR GOES BELOW
# 
hotel1_df$children <- as.integer(hotel1_df$children)
#hotel1_df$company <- as.integer(hotel1_df$company)
hotel1_df$agent <- as.integer(hotel1_df$agent)

str(hotel1_df)
```


## 5. Solution Implementation

### 5.a) Questions

```{r}
# 1. How many bookings were cancelled?
# ---
# Visualisation: Barplot

hotel1_df %>% 
  group_by(hotel) %>%
  summarize(is_canceled = sum(is_canceled))

# Plotting the chart
ggplot(hotel1_df, aes(x = hotel, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Total Hotel Cancellations", 
    x = "Hotel", 
    y = "Total Cancellations"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


```{r}
# 2. What was the booking ratio between resort hotel and city hotel?
# ---
# Barplot of booking ratio between resort hotel and city hotel

hotel1_df %>% 
  filter(hotel == "resort hotel",hotel == "city hotel") %>%
  group_by(hotel) %>%
  summarize(bookings = sum(previous_bookings_not_canceled))

# Plotting the chart
ggplot(hotel1_df, aes(x = hotel, y = previous_bookings_not_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Hotel Booking Ratio between resort and city hotels", 
    x = "Hotel", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


```{r}
# 3. What was the percentage of booking for each year?
# ---
# 
yearbookings <- hotel1_df %>%
  select(arrival_date_year,previous_bookings_not_canceled) %>%
  group_by(arrival_date_year)%>%
  summarize(previous_bookings_not_canceled = sum(previous_bookings_not_canceled))
  
yearbookings
```
```{r}
ggplot(data=yearbookings, aes(x="", y=previous_bookings_not_canceled, fill=as.factor(arrival_date_year))) +
  geom_bar(
    stat="identity", width=0.5, color = "white"
  ) +
  coord_polar("y") + 
  geom_text(
    aes(label = paste0(round(previous_bookings_not_canceled / sum(previous_bookings_not_canceled) * 100, 1), "%")), position = position_stack(vjust = 0.5)
  ) +
  scale_fill_brewer(palette="Set1") +   
  labs(
    fill = "arrival_date_year", 
    title = "Average reservations in hotel per year"
  ) +
  theme_void()
```

```{r}
# 4. Which were the most busiest months for hotels?
# ---
# 
hotelmonth_df <- hotel1_df %>%
  select(arrival_date_month, previous_bookings_not_canceled) %>%
    arrange(desc(previous_bookings_not_canceled)) %>%
  group_by(arrival_date_month)%>%
  summarize(previous_bookings_not_canceled = sum(previous_bookings_not_canceled))

hotelmonth_df
# Plotting
ggplot(hotelmonth_df, aes(x = arrival_date_month, y = previous_bookings_not_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  )  +
  labs(
    title = "Total Bookings per Month", 
    x = "Months", 
    y = "Total Booking"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  coord_flip()
```


```{r}
# 5. From which top 3 countries did most guests come from?
# --- 
# YOUR GOES BELOW
#
head(hotel1_df)

countryguest_df <- hotel1_df %>%
  select(country, adults,children,babies) %>%
  group_by(country)%>%
  summarize(Total_Guests = sum(adults+children+babies))%>%
  arrange(desc(Total_Guests))
head(countryguest_df,3)
```
```{r}

```

```{r}
# 6.a) How long do most people stay in hotels?
#   b) By city and resort? Separate the data by hotel
# ---
# 


gueststay_df <- hotel1_df %>%
  select(hotel, adults,stays_in_weekend_nights,stays_in_week_nights) %>%
  group_by(hotel)%>%
  summarize(Total_Stay = round(mean(stays_in_weekend_nights+stays_in_week_nights)))
  
gueststay_df
```


```{r}
# 7. Which was the most booked accommodation type (Single, Couple, Family)?
# ---
#Assumption
#if adults is 1 and no babies or children and booking 
accomodation_df <- hotel1_df %>%
  select(reserved_room_type, previous_bookings_not_canceled) %>%
  group_by(reserved_room_type)%>%
  summarize(Total_Bookings = sum(previous_bookings_not_canceled))%>%
  arrange(desc(Total_Bookings))
accomodation_df
```

### 5.b) Recommendations
From the above analysis, below are our recommendations:
  - people tend to stay longer in resorts than city hotels
  - highest number of bookings are done in may while the least are done in August

## 6. Challenging your Solution
In this step, we review our solution and implement approaches that could potentially provide a better outcome. In our case, we could propose the following question that wasn't answered in our solution because it couldn't have greatly contributed to our recommendation.

```{r}
# When should hotels provide special offers?
# --- this should be provided in the months between August and December where we see decline in bookings, this would motivate clients to book more.

# YOUR GOES BELOW
# 

hotelmonth_df <- hotel1_df %>%
  select(arrival_date_month, previous_bookings_not_canceled) %>%
    arrange(desc(previous_bookings_not_canceled)) %>%
  group_by(arrival_date_month)%>%
  summarize(previous_bookings_not_canceled = sum(previous_bookings_not_canceled))

hotelmonth_df
# Plotting
ggplot(hotelmonth_df, aes(x = arrival_date_month, y = previous_bookings_not_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  )  +
  labs(
    title = "Total Bookings per Month", 
    x = "Months", 
    y = "Total Booking"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + 
  coord_flip()
```

Our observations:
-We have high booking between January and May with the climax in May then that reduces as we progress towards the last months of the year.

How does this observation tie to our solution?
- Hotels should provide special offers between August and December so as to boost the booking which tend to go down during those months

## 7. Follow up questions
During this step, you rethink and propose other ways that you can improve your solution.

a). Did we have the right data?
  the data was right for the better part of the analysys, however, on the last question on the most accomodation type done(Single ,Couple or Family) there was no sufficient data to make those deductions and one is left with alot of assumptions to make the data work.
  
b). Do we need other data to answer our question?
    yes we need more data especially to give more insights on the customer type and other factors that would contribute to hotel bookings
    
c). Did we have the right question?
  The questions were right to make the necessary deductions from the data provided.


